---
title: "R intro"
author: "Murray Cadzow"
date: "3 November 2015"
output: html_document
---

# Lesson 1


Covered in this lesson:

- what is R/RStudio

- getting help

- basic operations

- variables

- reading and writing data

- R scripts


## What is R/RStudio?

"R is a language and environment for statistical computing and graphics"

Commands are entered and run in the R commandline

RStudio is an interface for R

## Getting help

On the prompt you can type _??command_ or _help(command)_ and the documentation for the command will appear

You can also refer to the R manual https://cran.r-project.org/doc/manuals/r-release/R-intro.html

Or www.google.com with "R command" or "R thing I want to do"

## Basic operations

### Working directory

In R everything you do is in reference to what is know as your 'working directory'
This is the folder on your computer that R will read and write data to by default

You can find your current working directory with the command
```{r, eval=FALSE}
getwd()
```

likewise you can change your working directory with
```{r, eval=FALSE}
setwd(dir = "path/to/directory")
```
or in RStudio in the Session menu -> Set Working Driectory

### Operators
The basic arithmetic operatiors in R are:

- addition +

- subtraction -

- multiplication *

- division /

- parentheses ( )

- remainder division %%

- exponentiation ^

- assignment <-

Order of operations follows BEDMAS, with **assignment being evaluated last**

In your R prompt try the following:
```{r, eval = FALSE, echo=TRUE}
2 + 5
8 - 1
3 * 4
6 / 2
1 * ( 3 - 1 )
7 %% 6
2 ^ 3
a <- 5
```


## Variables

Variables are used in R to store things, these can be numbers, letters or even what other variables contain

It is good practice to name your variable so that it describes what it contains

Variables can be called anything you like so long as you follow these rules:
- must start with a letter

- cannot contain symbols other than underscore or period ( _ or .)

- numbers (except at start) and upper or lower case letters are allowed

- cannot be a reserved word


Examples of variable names

mydata

Mydata

myData

my_data



In the last example you used the assignment operator ( <- ) to store the value 5 in the variable called 'a'

To view what a variable contains you just enter the variable name into the R prompt
```{r, eval=TRUE, echo = FALSE}
a <- 5
```

```{r, eval=TRUE, echo = TRUE}
a
```

Variables can be overwritten by assigning something different to them
Variables only contain the last item assigned
```{r, eval = TRUE, echo = TRUE}
weight <- 64
height <- 168
weight
height
height <- height / 100
bmi <- weight / height ^ 2

```

What do weight, height, and bmi now contain?

```{r}
weight
height
bmi
```

## Reading and Writing data

### Reading

Just as we can store single values into variables, we can store entire datasets in a variable

Generally the data we want to read in will be in either tab or comma separated format

We have to specify the filename, if the file has a header at and the field seperating character ( \\t = tab )

If you don't assign the output from read.delim to a variable it will just ouput the entire file into your prompt
```{r, eval=FALSE, echo = TRUE}
# for tab separated data
mydata <- read.delim(file = "dummy_data.txt", header= TRUE, sep = "\t", stringsAsFactors = FALSE)

# for comma separated data
mydata <- read.delim(file = "dummy_data.csv", header = TRUE, sep = ",", stringsAsFactors = FALSE)
```

### Writing

Writing files is very similar with a few minor differences

```{r, eval = FALSE, echo = TRUE}
write.table(data = mydata, file = "mydata_out.txt", sep = "\t", col.names = TRUE, row.names= FALSE, quote = FALSE )
```


# Lesson 2

This lesson is on storing and accessing data in R
it will cover:

- viewing data

- data types

- data structures

- naming/indices

- gotchas

## Viewing data

First we need some data to look at
```{r, eval = TRUE, echo = TRUE}
mydata <- read.delim(file = "dummy_data.txt", header = TRUE, stringsAsFactors = FALSE, sep ="\t")
```

Lets look at the dimensions of this dataset
```{r, eval=FALSE, echo=TRUE}
dim(mydata)
```

This tells us how many rows and columns we are dealing with

We can also find out what data structure our data has been put into

```{r, eval = FALSE, echo = TRUE}
class(mydata)
```

We can see that our data has been put into a ```r class(mydata)``` structure. More on this later

To view the first entries of our data we use the _head()_ function
```{r, eval = FALSE, echo = TRUE}
head(mydata)
```

Similarily, we can view the last entries of our data we use the _tail()_ function
```{r, eval = FALSE, echo = TRUE}
tail(mydata)
```

In **RStudio** we can use the _View()_ function to open our data in a familiar spreadsheet view
```{r, eval=FALSE, echo = TRUE}
View(mydata)
```

## Data Types

In R there are different data types automatically assigned depending on your data
the main data types are:



- numeric

- integer

- logical

- character

- factor

- missing values


### Numeric/Integer

Numeric and integer data types are exactly what they sound like - numbers

e.g. 1, 3.6 100, 5e7, 2e-2

The integer data type has to be specifically created

You can test if you have a numeric data type by using either the _class()_ or _is.numeric()_ functions


### Logical

The logical data type consists of the values **TRUE** and **FALSE**, and is often created as the result of the comparison operators (>,<,>=,<=,!,!=, ==, |, &)




### Character

A character or string of characters in R needs to be quoted, this is so R can distinguish between variable names and character data

Characters are any number,letter, or symbol on the keyboard including the white space (tab "\\t", space " ", and newline "\\n")

Placing a number in quotes forces it to be a character data type

eg "a", "a string of characters", "%", "2"

You can test if you have a character data type by using either the _class()_ or _is.character()_ functions




### Factors

Factors are a special data type in R and are used for groups with a set of levels

Levels can be either numeric or characters but being a factor data type, the possible __levels__/groups are stored too

In the _read.delim()_ function earlier we specified stringsAsFactors = FALSE, this is to override the default settings where all strings are read in as the factor data type

To demonstrate what a factor looks like we'll change a character data type into a factor with the _factor()_ function

```{r}
factor( c("control", "control", "case", "control", "case"))

factor( c(1,1,5,3,3,3,2))
```

To find the levels for a factor data type use the _levels()_ function




### Missing values

Missing values in R are assigned a special datatype called **NA**

NA acts as a place holder but often causes errors in functions when not removed

R provides a few useful functions for dealing with NA
```{r, echo = TRUE, eval = TRUE}
missing <- NA
missing
is.na(missing)


mydata <- c(1, 2, NA, 3, 4)

# example error for missing data
sum(data = mydata)
```

You can see that if you have an **NA** in your data, you will get back **NA** as a result

We can either remove the NA prior to using it, or functions often have the option to temporarily ignore NAs using na.rm = TRUE

```{r, echo = TRUE, eval = TRUE}
mydata
sum( data = mydata, na.rm = TRUE )

mydata <- na.omit(mydata)
mydata
sum(data = mydata)
```


## Data Structures

In R we arrange multiple data into structures based on the content

These structures include:

- vectors

- matrices

- lists

- dataframes

The most common of these are vectors and dataframes

### Vectors

vectors are the most basic data structure in R and consist of 0 (empty vector) or more values of any data type

Vectors are created using the function _c()_ - short for concatenate
```{r, echo = TRUE, eval = TRUE}
# an empty vector
c()

# numeric vector
c(1,2,3,4,5)

# logical vector
c(TRUE, FALSE, TRUE)

# character vector
c("a", "b", "c")

# a multitype vector
c(1, "a", FALSE)
```

We can also find out how many items are in a vector with the _length()_ function
```{r, echo = TRUE, eval = TRUE}
mydata <- c(1,5,3,4,2)
length(mydata)
```

### Matrices

A matrix is a 2 dimensional rectangular layout of data elements of a single data type

```{r, echo = TRUE, eval = TRUE}
mat <- matrix( data = c(1,2,3,4,5,6), nrow = 2, ncol = 3, byrow = TRUE)
mat
```

### Lists

A list consists of 0 or more elements, each element can be of same or differing lengths as well as differing data type and/or structure

```{r, eval = TRUE, echo = TRUE}
# empty list
list()

# 2 vectors
list1 <- list( c(1,2,3), c('a', 'b'))
list1

# vector and matrix
list2 <- list(mydata, mat)
list2
```

### Dataframes

Dataframes can be thought of as similar to a spreadsheet, made up of vectors forming columns of the same length. The number length of the vectors is the number of rows
```{r, eval = TRUE, echo = TRUE}
height <- c(1.5, 1.6, 1.55)
weight <- c(60,63,62)

df <- data.frame(height, weight )
df

# number rows
nrow(df)
# number of columns
ncol(df)

# dimensions (rows, cols)
dim(df)
```

## Naming/Indices


### Naming

We can find out different attributes from our dataframe such as dimensions
```{r, eval = TRUE, echo = TRUE}


# names of columns
names(df)

# column names
colnames(df)

# row names
rownames(df)
```

You can also change the names, row names, column names

The replacement vector has to match the length of the current names vector

```{r, eval = TRUE, echo = TRUE}
names(df) <- c("height_m", "weight_kg")
names(df)

rownames(df) <- c("row1", "row2", "row3")
rownames(df)
```


You can also use these column names to access only that column

To do this we use the variable name followed by the "$" symbol followed by the column name
```{r, echo = TRUE, eval = TRUE}
df$height_m
```

### Indices

In R, each item can also be refered to by its position (i) using the square brackets [i]

Vectors only have a single index, whereas matrices and dataframes have 2 in the form [row, column]

Lists are special in that [[i]] is used to access the list element


```{r, echo = TRUE, eval = TRUE}
dim(df)
df

# first row, first column
df[1,1]

# third row, first column
df[3,1]

# first and second rows, second column
df[c(1,2), 2]
# example error: second row, third column
df[2,3]
```

If you leave a index blank then R gives you everything
```{r, eval=TRUE, echo=TRUE}
# all of row 1
df[1, ]

# all of column 1
df[, 1]

# everything
df[ , ]
```

R also has a special feature for generating a vector of sequences of numbers with a:b

We can use this feature to select multiple sequencial items

```{r, eval = TRUE, echo = TRUE}
1:3

5:9

# select rows 1 to 3, column 2
df[1:3, 2]
```

## Gotchas

- blanks not becoming NAs

- auto factorisation when reading data in

- NAs being used by a function

- data not being the expected type by function

- index doesn't exist


# Lesson 3

In this lesson we will cover subsetting your data using:

- logical operations

- subset()

- square brackets

## Logical operations

Before we can subset we need to know how to create the conditions for R to know what to select

We do this through logical comparisions

The logical operators as mentioned before are:

- == ( equality )

- != ( not equal)

- \> (greater than)

- \>= (greater than or equal to)

- < (less than)

- <= (less than or equal to)

- ! (not)

- \| (or)

- & (and)

The result of using these logical comparisons is a vector of **TRUE**s and **FALSE**s

Here are some examples:
```{r, eval = TRUE, echo = TRUE}
df$weight
df$weight_kg < 62

# using not
!( df$weight_kg < 62)

# equality
df$weight_kg == 62

# not equality
df$weight_kg != 62

# combine conditions
df$weight > 62 | df$weight < 62

# different types
3 < 'a'
```

**NOTE** comparisons will be performed on the "lowest common data type" which is usually _character_ when different types are involved


## _Subset()_

The subset function provides a easy way to subset your data into groups based on your logical conditions

We'll use the dummy data for these examples

```{r, echo = TRUE, eval = TRUE}
# read data in
mydata <- read.delim(file = "dummy_data.txt", header = TRUE, stringsAsFactors = FALSE, sep ="\t")

# what does it look like?
dim(mydata)
head(mydata)

# select patients under 40
under40 <- subset(x = mydata, mydata$AGECOL < 40)

# how many people do we now have?
dim(under40)

# from our original data how many people are male (SEX == 1) and have gout (GOUTAFFSTAT == 2)
male_gout <- subset(x = mydata, mydata$SEX == 1 & mydata$GOUTAFFSTAT == 2)
dim(male_gout)

```

### Subset with square brackets

Just as we can use square brackets and indices to select items we want, instead of indices we can provide a logical condition


```{r, echo = TRUE, eval = TRUE}
# repeat under 40 selection
# here we're saying we want all the columns but select the rows where the condition is met
under40_repeat <- mydata[mydata$AGECOL < 40 , ]
dim(under40_repeat)

# who has missing age?
missing_age <- mydata[is.na(mydata$AGECOL),"PATIENT"]
missing_age
```


# Lesson 4

Data exploration