---
title: "R intro"
author: "Murray Cadzow"
date: "3 November 2015"
output: html_document
---

# Lesson 1

Covered in this lesson:

- what is R/RStudio

- basic operations

- variables

- reading and writing data

- R scripts


## What is R/RStudio?

"R is a language and environment for statistical computing and graphics"

Commands are entered and run in the R commandline

RStudio is an interface for R


## Basic operations

### Working directory

In R everything you do is in reference to what is know as your 'working directory'
This is the folder on your computer that R will read and write data to by default

You can find your current working directory with the command
```{r, eval=FALSE}
getwd()
```

likewise you can change your working directory with
```{r, eval=FALSE}
setwd(dir = "path/to/directory")
```
or in RStudio in the Session menu -> Set Working Driectory

### Operators
The basic arithmetic operatiors in R are:

- addition +

- subtraction -

- multiplication *

- division /

- parentheses ( )

- remainder division %%

- exponentiation ^

- assignment <-

Order of operations follows BEDMAS, with **assignment being evaluated last**

In your R prompt try the following:
```{r, eval = FALSE, echo=TRUE}
2 + 5
8 - 1
3 * 4
6 / 2
1 * ( 3 - 1 )
7 %% 6
2 ^ 3
a <- 5
```


## Variables

Variables are used in R to store things, these can be numbers, letters or even what other variables contain

It is good practice to name your variable so that it describes what it contains

Variables can be called anything you like so long as you follow these rules:
- must start with a letter

- cannot contain symbols other than underscore or period ( _ or .)

- numbers (except at start) and upper or lower case letters are allowed

- cannot be a reserved word


Examples of variable names

mydata

Mydata

myData

my_data



In the last example you used the assignment operator ( <- ) to store the value 5 in the variable called 'a'

To view what a variable contains you just enter the variable name into the R prompt
```{r, eval=TRUE, echo = FALSE}
a <- 5
```

```{r, eval=TRUE, echo = TRUE}
a
```

Variables can be overwritten by assigning something different to them
Variables only contain the last item assigned
```{r, eval = TRUE, echo = TRUE}
weight <- 64
height <- 168
weight
height
height <- height / 100
bmi <- weight / height ^ 2

```

What do weight, height, and bmi now contain?

```{r}
weight
height
bmi
```

## Reading and Writing data

### Reading

Just as we can store single values into variables, we can store entire datasets in a variable

Generally the data we want to read in will be in either tab or comma separated format

We have to specify the filename, if the file has a header at and the field seperating character ( \\t = tab )

If you don't assign the output from read.delim to a variable it will just ouput the entire file into your prompt
```{r, eval=FALSE, echo = TRUE}
# for tab separated data
mydata <- read.delim(file = "mydata.txt", header= TRUE, sep = "\t", stringsAsFactors = FALSE)

# for comma separated data
mydata <- read.delim(file = "mydata.txt", header = TRUE, sep = ",", stringsAsFactors = FALSE)
```

### Writing

Writing files is very similar with a few minor differences

```{r, eval = FALSE, echo = TRUE}
write.table(data = mydata, file = "mydata_out.txt", sep = "\t", col.names = TRUE, row.names= FALSE, quote = FALSE )
```


# Lesson 2

This lesson is on storing and accessing data in R
it will cover:

- viewing data

- data types

- data structures

- naming

- gotchas

## Viewing data

First we need some data to look at
```{r, eval = FALSE, echo = TRUE}
data <- read.delim(file = "file.txt", header = TRUE, stringsAsFactors = FALSE, sep ="\t")
```

Lets look at the dimensions of this dataset
```{r, eval=FALSE, echo=TRUE}
dim(mydata)
```

This tells us how many rows and columns we are dealing with

We can also find out what data structure our data has been put into

```{r, eval = FALSE, echo = TRUE}
class(mydata)
```

We can see that our data has been put into a ```r class(data)``` structure. More on this later


## Data Types

In R there are different data types automatically assigned depending on your data
the main data types are:

- numeric

- integer

- character

- factor

- boolean

- NA